---
title: "Reproducible Research"
author: "Luiz Bergo"
date: "16/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A glimpse at NDCD Storm Events in USA from 1950 - 2011

**DICLAIMER:** This document is prerequisite for conclusion of Reproducible Research course offered as part of Data Science path at Coursera. The author made efforts to bring a good analysis, even so, due to some lack on R knlowdge, course deadline and the difficult to treat the data set, errors may occured. The reader is invited to audit the steps and the analysis itself.

### What is this about

The main idea of this work is analyse a dataset from **Storms** occured at the USA in the period of 1.950 to 2011. The database is curated by **National Climatic Data Center (NCDC)** and most of the data were manually inputed leading to *typo* issues. To fix them some techniques of *data cleaning* where applied to focus on answer main questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

```{r libraries, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(taRifx)
library(lattice)
library(reshape2)
```

```{r cars}

```

### Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
