---
title: "Reproducible Research - Analysis on piece of storm data from NDCD (1950-2011)"
author: "Luiz Bergo"
date: "16/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A glimpse at NDCD Storm Events in USA from 1950 - 2011

**DICLAIMER:** This document is prerequisite for conclusion of Reproducible Research course offered as part of Data Science path at Coursera. The author made efforts to bring a good analysis, even so, due to some lack on R knlowdge, course deadline and the difficult to treat the data set, errors may occured. The reader is invited to audit the steps and the analysis itself.

### What is this about

The main idea of this work is analyse a dataset from **Storms** occured at the USA in the period of 1.950 to 2011. The database is curated by **National Climatic Data Center (NCDC)** and most of the data were manually inputed leading to *typo* issues. To fix them some techniques of *data cleaning* where applied to focus on answer main questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


### Setting up enviroment

The packages necesseary to analysis are:

- dplyr
- ggplot2
- gridExtra
- taRifx
- lattice
- reshape2

If not installed some code chunk may broke, so it´s recomended to have all of those packages installed prior continue.

```{r libraries, message=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(taRifx)
library(lattice)
library(reshape2)
```

The data set used in this work is a subset of the entire data that is availabe at NCDC servers. So, it´s assured the reproducibility if the dataset available at this link is used.

<https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>

It´s worthy to note that the dataset has data for a long time passed, I mean, 1950, most of that manually inputed, so the blank fields were filled with NA.

```{r download_dataset}

dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileZip <- "stormData.bz2"
dirData <- "./data"
fileCSV <- "stormData.csv"

if (!file.exists(fileZip)){
        download.file(dataURL,fileZip)
}

```

```{r cache=TRUE}

st_data <- read.csv("stormData.bz2", na.strings =c("","NA"), stringsAsFactors = FALSE)
st_data2 <- st_data # Creating a new dataset to avoid destroy original one.

```

```{r list_of_event_type, message=FALSE}
# This is a type list from oficial document of NDCC

eventlist <- c("Astronomical Low Tide","Avalanche","Blizzard","Coastal Flood","Cold/Wind Chill","Debris Flow","Dense Fog","Dense Smoke","Drought","Dust Devil","Dust Storm", "Excessive Heat","Extreme Cold/Wind Chill","Flash Flood","Flood","Freezing Fog","Frost/Freeze","Funnel Cloud","Hail","Heat","Heavy Rain","Heavy Snow","High Surf","High Wind","Hurricane/Typhoon","Ice Storm","Lakeshore Flood","Lake-Effect Snow","Lightning","Marine Hail","Marine High Wind","Marine Strong Wind","Marine Thunderstorm Wind","Rip Current","Seiche","Sleet","Storm Tide","Strong Wind","Thunderstorm Wind","Tornado","Tropical Depression","Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire","Winter Storm","Winter Weather")%>%
  toupper()

```


In a first glance at dataset we observe **37** variables with **902297** observations:

```{r dataview}

dim(st_data)
head(st_data)

```

Most of the variables won´t be necessary at this work.The key features to answers the question in fact may be reach observing the following variables:

- *EVTYPE*: classify the **type of storm**.
- *FATALITIES* and *INJURIES*: 
- *CROPDMG* and *PROPDMG*: respectively quantify the economic loss at **crops**. and **properties**.

For **Reproducible Research** Coursera practitioners, useful information can be found at <https://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/38y35MMiEeiERhLphT2-QA>.

Other useful information for a deep understand on datastructure can be found in the following link:

- NDCC data dictionary: <https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf> 
- NDCC data FAQ: <https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf>


### Data preparation

For purpose of this work the data preparation can be divided in three parts.

#### 1. Event typo treatment

As previous commented, the EVTYPE is a key to understand what type of impact brings more loss. The problem that raise is that this information is very noise due many factors from those we highlight:

1. Wrong typing.
2. Terminology rules and conecpts changes over the time.

The evidence of such explanation is the ***985*** unique values found in the **EVTYPE** column. 

A quick inspection on that is enough to percept many upper and lower issues, missplaced information like "summary of (...)"

```{r unique_original_EVTYPE}

length(unique(st_data$EVTYPE))

```

The EVTYPE where processed as follow:

```{r EVTYPE_preparation}

evtype <- st_data2$EVTYPE

evtype <- gsub("^ *","", st_data2$EVTYPE)%>% # Remove spaces from begin
  toupper() # Define events as upper case
evtype <- gsub("^SUMMARY",NA, evtype)  # Remove wrong entries

st_data2$EVTYPE_Treated <- NA

for (event in eventlist){
  val_exp <- paste("*",event,"*", sep="")
  filtered_events <- grep(val_exp,evtype)
  st_data2$EVTYPE_Treated[filtered_events]<- event
}

# Even after a first treatment there are 226.202 TSTM events distributed in 26 
# categories!!!!
# Routine to treat TSTM events
filtro_TSTM <- grep("*TSTM*", st_data2$EVTYPE)
st_data2$EVTYPE_Treated[filtro_TSTM]<- "THUNDERSTORM WIND"


```

The result of data treatment on ***EVTYPE*** is that the unique values are now  ***42*** and available at a new column named ***EVTYPE_Treated".

```{r}
length(unique(st_data2$EVTYPE_Treated))
```


#### 2. Injuries and fatalities

The information about health losses is splited in at two columns ***INJURIEs*** and ***FATALITIES***. To work with a small piece of data, this columns and ***EVTYPE_Treated*** where selected from prepared dataset. 

To prepared this data for plot the columns ***INJURIEs*** and ***FATALITIES*** were melted.

For a better analysis, the data the health was quantiled to evaluate the values corresponding to 80% of total of cases.

```{r preparation_health_data}

injurydata <- st_data2[,c("EVTYPE_Treated","FATALITIES","INJURIES")]
names(injurydata) <-c("Event", "TypeDamage","Count")

melt_injury <- melt(injurydata, variable.name = "TypeDamage", 
                  value.name = "Count")

gr_melt_injury <- group_by(melt_injury,Event)
sum_CountInjury <- summarize(gr_melt_injury, sum(Count))
names(sum_CountInjury) <-c("Event","Count")
na_event <- is.na(sum_CountInjury$Event) 
sum_CountInjury[na_event,"Event"] <- "MISSING VALUE (NA)"

# Filtering the cases that correspond to 80% of total injury and fatalities. 

cut_injury <- quantile(sum_CountInjury$Count, probs = c(0, 0.8, 1))
rl_injury <- filter(sum_CountInjury, Count > cut_injury [2])

```


#### 3. Crop and property

The crop and property losses cost were prepared in two steps:

- First:

The cost related to damage on properties and crops are available at columns ***CROPDMG*** and **PROPDMG*** but they are quite afected by ***CROPDMGEXP*** and ***PROPDMGEXP*** respectively.

The ***CROPDMGEXP*** and ***PROPDMGEXP*** are multipliers that vary from 1 to 1E9 and must be applyed to raw values. A reasonable explanation of this particularity in dataset is presented here  <https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html>.

The final result of this process is the creation of new the columns ***MULTICROP*** and ***MULTIPROP***.

```{r }

# Prevent typing problem with classifiers

st_data2$PROPDMGEXP <- toupper(st_data2$PROPDMGEXP)
st_data2$CROPDMGEXP <- toupper(st_data2$CROPDMGEXP)



labels_mult <- sort(unique(c(st_data2$PROPDMGEXP,st_data2$CROPDMGEXP)))
values_mult <- c(1,1,1,rep(10,9),1E9,100,1E3,1E6)


mult_props <- as.data.frame(cbind(labels_mult,values_mult)) %>%
  remove.factors()%>%
  rbind(c(NA,1,1))

names(mult_props) <- c("labels_mult","values_multprops")

mult_crops <- mult_props
names(mult_crops) <- c("labels_mult","values_multcrops")

# Transforming data set to include new columns

st_data2 <- merge(x = st_data2, y=mult_props, by.x = "PROPDMGEXP", 
                by.y = "labels_mult")

st_data2 <- merge(x = st_data2, y=mult_crops, by.x = "CROPDMGEXP", 
                  by.y = "labels_mult")%>%
  transform(values_multcrops = as.numeric(values_multcrops))%>%
  transform(values_multprops = as.numeric(values_multprops))%>%
  transform(PROPDMGCOST=PROPDMG*values_multprops)%>%
  transform(CROPDMGCOST=CROPDMG*values_multcrops)

st_data2$values_multcrops <- as.numeric(st_data2$values_multcrops)
st_data2$values_multprops <- as.numeric(st_data2$values_multprops)

#Applying the multpliers to raw data.

st_data2 <- transform(st_data2,MULTIPROP=PROPDMG*values_multprops)%>%
  transform(MULTICROP=CROPDMG*values_multcrops)

head(st_data2[,c("EVTYPE","MULTICROP","MULTIPROP")])
```


- Second:

The data related to costs were subset from altered dataset and molten to be prepared for plot.

For a better analysis, the loss data was quantiled to evaluate the values corresponding to 80% of total of losses cost.

```{r}

# Subseting cost data

costdata <- st_data2[,c("EVTYPE_Treated","PROPDMGCOST","CROPDMGCOST")]
names(costdata) <-c("Event", "TypeDamage","Value")

# Melting data

melt_cost <- melt(costdata, variable.name = "TypeDamage", 
                  value.name = "DMGValue")

gr_melt_cost <- group_by(melt_cost,Event)
sum_costDMG <- summarize(gr_melt_cost, sum(DMGValue))
names(sum_costDMG) <-c("Event","DMGValue")
na_event <- is.na(sum_costDMG$Event)
sum_costDMG[na_event,"Event"] <- "MISSING VALUE (NA)"

cut_event <- quantile(sum_costDMG$DMGValue, probs = c(0, 0.8, 1))
rl_event <- filter(sum_costDMG, DMGValue > cut_event[2])

```

## Results

Let me recap the two questions that drive this work:

***1. Across the United States, which types of events are most harmful with respect to population health?***

***2. Across the United States, which types of events have the greatest economic consequences?***

To answer the first question we present the graphic as follow.

```{r health_issues}
damage_injury <- ggplot(rl_injury, aes(x=reorder(Event,Count), y= Count))
damage_injury <- damage_injury  + geom_col()
damage_injury <- damage_injury  + coord_flip()
damage_injury <- damage_injury  + labs(title = "Health Losses by Weather Events in USA 1950 - 2011")
damage_injury <- damage_injury  + theme(plot.title = element_text(hjust = 0.5))
damage_injury <- damage_injury  + labs(y = "Main Injury and Death" , x = "Event type")
print(damage_injury)
```

After analysis it´s clear that ***TORNADO*** events are, far away from the others, the main cause of injuries and deaths due to storms events. We believe that extra effort and analysis must be done to understand how and in wich circustances a tornado becomes more or less harmful. The suggestions are:

- ***Extend the search*** including other databases and features to modelling tornado behaviors.
- Invest on communication channel to ***inform probability of tornado events*** within weeks ahead.
- Fund projects for shelter and other infrastructure to ***support population during tornado events***.

It´s worthy to note that the second and fourth cause is related to ***HEAT*** and ***FLOOD*** and this suggest that a problem may is related to overall  climate change, a big issue and a world class problem.

To answer the second question we present the graphic as follow.

```{r economic_losses}
damage_plot <- ggplot(rl_event, aes(x=reorder(Event,DMGValue), y= DMGValue))
damage_plot <- damage_plot  + geom_col()
damage_plot <- damage_plot  + coord_flip()
damage_plot <- damage_plot  + labs(title = "Economic Losses by Weather Events in USA 1950 - 2011")
damage_plot <- damage_plot  + theme(plot.title = element_text(hjust = 0.5))
damage_plot <- damage_plot  + labs(y = "Main losses (US$)" , x = "Event type")
print(damage_plot)
```

The first thing it´s worthy to note is the amount of "Missing Values" found in data set but with some property or crop damage related. At this time of problem investigation it´s not possible to explain why this happen, and some questions raised:

- Shall property or crop losses related to broken communication or other side effects that difficult a proper classification?
- Shall some property or crop losses informed as a fraud?

Aside the missing values three events are highlighted, ***FLOOD***, ***HURRICANE/TYPHOON***, and ***TORNADO***. 

In the previous session the suggestion to investigate ***TORNADO*** events may be apllied here too but aiming to investigate how such events affect crops and propperties, in the context of economic losses, this investigation should include ***HURRICANE/TORNADO***.

It is not surprising that ***FLOOD*** are the main reason for economic losses and is at the same time a question offen related to climate change. The suggestion to minimize effect of flood are:

1. Investigate areas with more ***probability of flood events***.
2. Maneauver crops and properties in the ***danger zones**.
3. Invest on development of new bulding technology with more harmony to the environment.


